---
title: "Controlling via residualising"
author:
  - name: Rogier Kievit
date: 06-01-2023
output:
  distill::distill_article:
    self_contained: false
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating blogs with Distill at:
# https://rstudio.github.io/distill/blog.html

```

this may or may not be useful - often reviewers might ask you to control for a certain variable. You can do this within the model, but another way is to residualize a variable for a covariate beforehand. In this toy example, you have height and weight, and perhaps you want to use one variable without the potential confounding of the other. With this v basic script you can see how to do it, and also how to use these new variables in plotting  (e.g. by showing who is taller/heavier than expected given a model)

it’s pretty basic but rarely illustrated (as far as I’ve seen) so may be worthwhile as an example, as I find myself showing this to many and varied people (in fact I just found an old script where I did almost the same :) )

```{r}
#how to adjust (residualize( a variable for a covariate before using the variable in other analyses

#creating fake data example with height and weight
height<-rnorm(100,165,10)
weight<-.4*height+rnorm(100,0,10)
plot(height,weight)

#modest correlation
cor(height,weight)

#fitting a linear model to predict weight
pred_weight<-lm(weight~height)
#saving the model
summary(pred_weight)

#saving the residuals - so weight adjusted for height differences
weight_residualised_for_height<-(pred_weight$residuals)
#this residualized weight still correlates with weight, as it should - people who are heavier than expected for their height
#will generally also be heavier full stop
cor(weight_residualised_for_height,weight)
#but not with height - so this is now a new kind of height, uncorrelated with weight
cor(weight_residualised_for_height,height)

#opposite model, predicting height from weight
pred_height<-lm(height~weight)
summary(pred_height)
#residual height: who is not just tall, but tall given their weight
height_residualised_for_weight<-(pred_height$residuals)
cor(height_residualised_for_weight,height)

#saving the four variables
all_data<-data.frame(height,weight,weight_residualised_for_height,height_residualised_for_weight)
#renaming
colnames(all_data)<-c('height','weight','resid_weight','resid_height')

#rescaling for plotting purposes
library(scales)
all_data$resid_weight<-rescale(all_data$resid_weight)
all_data$resid_height<-rescale(all_data$resid_height)

library(ggplot2)

#you can now plot people and plot the dots by size or colour to show who is heavier than expected given their height, or vice versa
#lighter blue, or bigger dots, mean people are heavier than expected given their weight, which may be a 
#'better' type of weight for certain purposes
ggplot(all_data,aes(height,weight))+geom_point(aes(size=resid_weight))+geom_smooth(method='lm')
ggplot(all_data,aes(height,weight))+geom_point(aes(col=resid_weight))+geom_smooth(method='lm')

#wrong residual so doesn't work
ggplot(all_data,aes(height,weight))+geom_point(aes(col=resid_height))+geom_smooth(method='lm')

#opposite association
ggplot(all_data,aes(weight,height))+geom_point(aes(size=resid_height))+geom_smooth(method='lm')
ggplot(all_data,aes(weight,height))+geom_point(aes(col=resid_height))+geom_smooth(method='lm')

#in future analysis, if you want to look at, say, the effect of height on health outcomes
#and someone says 'but height is confounded by weight', you now have a new height variable that we've forced to be uncorrelayed to weight, and vice versa

```

*Author: Rogier Kievit*
